---
title: "On The Role of Context in Reading Time Prediction"
output: html_notebook
---

```{r}
shhh <- suppressPackageStartupMessages # It's a library, so shhh!

shhh(library( mgcv ))
shhh(library(dplyr))
shhh(library(ggplot2))
shhh(library(lme4))
shhh(library(tidymv))
shhh(library(gamlss))
shhh(library(gsubfn))
shhh(library(lmerTest))
shhh(library(tidyverse))
shhh(library(boot))
shhh(library(rsample))
shhh(library(plotrix))
shhh(library(ggrepel))
shhh(library(mgcv))
shhh(library(relaimpo))
library(jmuOutlier) # For paired permutation tests

theme_set(theme_bw())
options(digits=4)
options(dplyr.summarise.inform = FALSE)
```

```{r}

set.seed(444)
langs = c("du", "en", "fi", "ge", "gr", "he", "it", "ko", "sp", "tr", "ru")
psychometrics = c("total_rt", "gaze_rt", "firstfix_rt")
models = c("mgpt_lc")
comps = c("target", "baseline")

```


```{r}

combined_df = data.frame()

for(lang in langs){
  
   df = read.csv(paste0("./merged_data/", lang, ".csv")) %>%
    filter(freq > 0, prev_freq > 0, prev2_freq > 0) %>%
    mutate( freq = -log2(freq), prev_freq = -log2(prev_freq), prev2_freq = -log2(prev2_freq),
            pmi = freq - surp, 
            prev_pmi = prev_freq - prev_surp, 
            prev2_pmi = prev2_freq - prev2_surp)
   
   combined_df = rbind(combined_df, df %>% mutate(lang = lang))
  
}
```

## EXPERIMENT 1: Linear models to interpret effects of context

```{r}

get_model_coeffs = function(form, df, mixed_effects, num_folds=10){
  
  folds <- cut(seq(1,nrow(df)),breaks=num_folds,labels=FALSE)
  
  coeffs = data.frame()
  for(i in 1:num_folds){
    testIndexes = which(folds==i,arr.ind=TRUE)
    trainData = df[-testIndexes,]
    
    trainData = trainData %>% 
      mutate(freq = (freq - mean(freq))/sd(freq), prev_freq = (prev_freq - mean(prev_freq))/sd(prev_freq), prev2_freq = (prev2_freq - mean(prev2_freq))/sd(prev2_freq),
           surp = (surp - mean(surp))/sd(surp), prev_surp = (prev_surp - mean(prev_surp))/sd(prev_surp), prev2_surp = (prev2_surp - mean(prev2_surp))/sd(prev2_surp),
           pmi = (pmi - mean(pmi))/sd(pmi), prev_pmi = (prev_pmi - mean(prev_pmi))/sd(prev_pmi), prev2_pmi = (prev2_pmi - mean(prev2_pmi))/sd(prev2_pmi)) %>%
      mutate(orthogonal_surp = surp - (cov(trainData$freq, trainData$surp)/cov(trainData$freq, trainData$freq)) * freq,
                                     prev_orthogonal_surp = prev_surp - (cov(trainData$prev_freq, trainData$prev_surp)/cov(trainData$prev_freq, trainData$prev_freq)) * prev_freq,
                                     orthogonal_pmi = pmi - (cov(trainData$freq, trainData$pmi)/cov(trainData$freq, trainData$freq)) * freq,
                                     prev_orthogonal_pmi = prev_pmi - (cov(trainData$prev_freq, trainData$prev_pmi)/cov(trainData$prev_freq, trainData$prev_freq)) * prev_freq)

    if(mixed_effects){
      model = lmer(as.formula(form), data = trainData)
    } else {
      model = lm(as.formula(form), data = trainData)
    }

    coeffs = rbind(coeffs, model$coefficients)
  }
  #print(model$coefficients)
  return(coeffs)
}

get_model_importance = function(form, df, mixed_effects, num_folds=10){
  
  folds <- cut(seq(1,nrow(df)),breaks=num_folds,labels=FALSE)
  
  metrics = data.frame()
  for(i in 1:num_folds){
    testIndexes = which(folds==i,arr.ind=TRUE)
    trainData = df[-testIndexes,]
    
    trainData = trainData %>% 
      mutate(freq = freq - mean(freq), prev_freq = prev_freq - mean(prev_freq), prev2_freq = prev2_freq - mean(prev2_freq),
           surp = surp - mean(surp), prev_surp = prev_surp - mean(prev_surp), prev2_surp = prev2_surp - mean(prev2_surp),
           pmi = pmi - mean(pmi), prev_pmi = prev_pmi - mean(prev_pmi), prev2_pmi = prev2_pmi - mean(prev2_pmi),
           len = len - mean(len), prev_len = prev_len - mean(prev_len), prev2_len = prev2_len - mean(prev2_len)) %>%
      mutate(orthogonal_surp = surp - (cov(trainData$freq, trainData$surp)/cov(trainData$freq, trainData$freq)) * freq,
              prev_orthogonal_surp = prev_surp - (cov(trainData$prev_freq, trainData$prev_surp)/cov(trainData$prev_freq, trainData$prev_freq)) * prev_freq,
              orthogonal_pmi = pmi - (cov(trainData$freq, trainData$pmi)/cov(trainData$freq, trainData$freq)) * freq,
              prev_orthogonal_pmi = prev_pmi - (cov(trainData$prev_freq, trainData$prev_pmi)/cov(trainData$prev_freq, trainData$prev_freq)) * prev_freq,
              orthogonal_len = len - (cov(trainData$freq, trainData$len)/cov(trainData$freq, trainData$freq)) * freq,
              prev_orthogonal_len = prev_len - (cov(trainData$prev_freq, trainData$prev_len)/cov(trainData$prev_freq, trainData$prev_freq)) * prev_freq)

    if(mixed_effects){
      model = lmer(as.formula(form), data = trainData)
    } else {
      model = lm(as.formula(form), data = trainData)
    }
    
    res = calc.relimp(model, type = c("lmg"))
    metrics = rbind(metrics, c(R2=res$R2, res$lmg))
  }
  
  
  return(metrics)
}

```


```{r}

toggle_surp = FALSE
#toggle_surp = TRUE

xlang_coeff_df = data.frame(matrix(ncol = 5, nrow = 0))
columnnames = c("intercept", "surp OR pmi", "previous surp OR pmi", "frequency", "previous frequency")
colnames(xlang_coeff_df) = columnnames

xlang_lmg_df = data.frame(matrix(ncol = 5, nrow = 0))
columnnames2 = c("R2", "surp OR pmi", "previous surp OR pmi", "frequency", "previous frequency")
colnames(xlang_lmg_df) = columnnames2

if (toggle_surp) {
  regression_form = "gaze_rt ~ surp + prev_surp + freq + prev_freq"
  #regression_form = "gaze_rt ~ surp + freq"
} else {
  regression_form = "gaze_rt ~ pmi + prev_pmi + freq + prev_freq"
}


for (l in langs) {
  
  print(paste0("Fitting standard model for ", l))
  
  df = read.csv(paste0("./merged_data/", l, ".csv")) %>%
    filter(freq > 0, prev_freq > 0, prev2_freq > 0) %>%
    mutate( freq = -log2(freq), prev_freq = -log2(prev_freq), prev2_freq = -log2(prev2_freq),
            pmi = freq - surp, 
            prev_pmi = prev_freq - prev_surp, 
            prev2_pmi = prev2_freq - prev2_surp)
  
  for (m in models) {
    
    df_eval = df %>% filter(model == m) %>% drop_na()
      
    coeff_df = get_model_coeffs(regression_form, df=df, mixed_effects = F)
    colnames(coeff_df) = columnnames
    
    lmg_df = get_model_importance(regression_form, df=df, mixed_effects = F)
    colnames(lmg_df) = columnnames2
    
    if (toggle_surp) {
      xlang_coeff_df = rbind(xlang_coeff_df, coeff_df %>% mutate(lang = l, model = m, orthogonalized = FALSE, surp = TRUE))
      xlang_lmg_df = rbind(xlang_lmg_df, lmg_df %>% mutate(lang = l, model = m, orthogonalized = FALSE, surp = TRUE))
    } else{
      xlang_coeff_df = rbind(xlang_coeff_df, coeff_df %>% mutate(lang = l, model = m, orthogonalized = FALSE, surp = FALSE))
      xlang_lmg_df = rbind(xlang_lmg_df, lmg_df %>% mutate(lang = l, model = m, orthogonalized = FALSE, surp = FALSE))
    }
      

  }
}

if (toggle_surp) {
  regression_form = "gaze_rt ~ orthogonal_surp + prev_orthogonal_surp + freq + prev_freq"
  #regression_form = "gaze_rt ~ orthogonal_surp + freq"
} else {
  regression_form = "gaze_rt ~ orthogonal_pmi + prev_orthogonal_pmi + freq + prev_freq"
}


for (l in langs) {
  
  print(paste0("Fitting orthogonalized model for ", l))
  
  df = read.csv(paste0("./merged_data/", l, ".csv")) %>%
    filter(freq > 0, prev_freq > 0, prev2_freq > 0) %>%
    mutate( freq = -log2(freq), prev_freq = -log2(prev_freq), prev2_freq = -log2(prev2_freq),
            pmi = freq - surp, 
            prev_pmi = prev_freq - prev_surp, 
            prev2_pmi = prev2_freq - prev2_surp)
  
  for (m in models) {
    
    df_eval = df %>% filter(model == m) %>% drop_na()
      
    coeff_df = get_model_coeffs(regression_form, df=df, mixed_effects = F)
    colnames(coeff_df) = columnnames
    
    lmg_df = get_model_importance(regression_form, df=df, mixed_effects = F)
    colnames(lmg_df) = columnnames2
    
    if (toggle_surp) {
      xlang_coeff_df = rbind(xlang_coeff_df, coeff_df %>% mutate(lang = l, model = m, orthogonalized = TRUE, surp = TRUE))
      xlang_lmg_df = rbind(xlang_lmg_df, lmg_df %>% mutate(lang = l, model = m, orthogonalized = TRUE, surp = TRUE))
    } else{
      xlang_coeff_df = rbind(xlang_coeff_df, coeff_df %>% mutate(lang = l, model = m, orthogonalized = TRUE, surp = FALSE))
      xlang_lmg_df = rbind(xlang_lmg_df, lmg_df %>% mutate(lang = l, model = m, orthogonalized = TRUE, surp = FALSE))
    }
      

  }
}

if (toggle_surp) {
      xlang_coeff_df_surp = xlang_coeff_df
      xlang_lmg_df_surp = xlang_lmg_df
    } else {
      xlang_coeff_df_pmi = xlang_coeff_df
      xlang_lmg_df_pmi = xlang_lmg_df
}

```

## IMPORTANT: Need to run the previous block for both values of toggle_surp before proceeding

```{r}

xlang_coeff_df = rbind(xlang_coeff_df_surp, xlang_coeff_df_pmi)
xlang_lmg_df = rbind(xlang_lmg_df_surp, xlang_lmg_df_pmi)

```


## Plot the LMG values

```{r}

coeff_plotting_df = xlang_lmg_df %>%
  gather(name, value, c(2:length(columnnames2))) %>%
  group_by(lang, name, model, orthogonalized, surp) %>%
    summarise(m = mean(value),
              s = std.error(value),
              upper = m + s * 1.96,
              lower = m - s * 1.96) %>%
  ungroup() %>%
  filter((surp == TRUE & orthogonalized == TRUE) | (surp == TRUE & orthogonalized == FALSE) | (surp == FALSE & orthogonalized == FALSE) ) %>%
  mutate(position = case_when(
    #name %in% c("R2") ~ "0",
    name %in% c("frequency", "surp OR pmi") ~ "0",
    name %in% c("previous frequency", "previous surp OR pmi") ~ "1",
  )) %>%
  mutate(lmodel = case_when(
    orthogonalized == FALSE & surp == TRUE ~ "Surprisal",
    orthogonalized == FALSE & surp == FALSE ~ "PMI",
    orthogonalized == TRUE & surp == TRUE ~ "Orth. Surprisal",
  )) %>%
  mutate(predictor = case_when(
    name %in% c("frequency", "previous frequency") ~ "Frequency",
    name %in% c("surp OR pmi", "previous surp OR pmi") ~ "Context",
    #name %in% c("R2") ~ "R2",
  )) %>%
  mutate(lmodel = factor(lmodel, levels = c("Surprisal", "PMI", "Orth. Surprisal"))) %>%
  mutate(predictor = factor(predictor, levels = c("Frequency", "Context"), labels = c("Frequency", "Context (Surprisal / PMI / Orthogonalized Surprisal)"))) %>%
  mutate(position = factor(position, labels = c("w t", "w t-1"), levels = c("0", "1")))
```

```{r}
coeff_plotting_df %>%
    mutate(lang = factor(lang, levels = c("du", "en", "fi", "ge", "gr", "he", "it", "ko", "ru", "sp", "tr"),
       labels = c("Dutch", "English", "Finnish", "German", "Greek", "Hebrew", "Italian", "Korean", "Russian",
                  "Spanish", "Turkish"))) %>%
  ggplot(aes(x = position, y = m, fill = predictor)) +
    geom_bar(stat="identity", position = position_dodge(width = 0.9)) +
    geom_errorbar(aes(ymin=lower, ymax=upper), position = position_dodge(width = 0.9), width = 0.1) +
    geom_hline(yintercept = 0, color = "black") +
  ylab("Explained Variance (LMG)") +
  facet_grid(lmodel~lang, scales = "free_y") +
  scale_x_discrete(labels = c(bquote(w[t]), bquote(w[t-1]))) +
  #scale_color_manual(values = c("#225ea8", "#41b6c4", "#a1dab4")) +
  theme(
    text = element_text(family = "serif", size=14),
    axis.title.x = element_blank(),
    legend.position = "bottom"
  )

ggsave("./images/orthogonalized_comparison_lmg.pdf", width = 9, height = 5)

```

#### Doing the same analysis with orthogonalized length

```{r}

#toggle_surp = FALSE
toggle_surp = TRUE

xlang_lmg_df = data.frame(matrix(ncol = 7, nrow = 0))
columnnames = c("R2", "surp OR pmi", "previous surp OR pmi", "frequency", "previous frequency", "length", "previous length")
colnames(xlang_lmg_df) = columnnames


if (toggle_surp) {
  regression_form = "gaze_rt ~ surp + prev_surp + freq + prev_freq + len + prev_len"
} else {
  regression_form = "gaze_rt ~ pmi + prev_pmi + freq + prev_freq + len + prev_len"
}

for (l in langs) {
  
  print(paste0("Fitting standard model for ", l))
  
  df = read.csv(paste0("./merged_data/", l, ".csv")) %>%
    filter(freq > 0, prev_freq > 0, prev2_freq > 0) %>%
    mutate( freq = -log2(freq), prev_freq = -log2(prev_freq), prev2_freq = -log2(prev2_freq),
            pmi = freq - surp, 
            prev_pmi = prev_freq - prev_surp, 
            prev2_pmi = prev2_freq - prev2_surp)
  
  for (m in models) {
    
    df_eval = df %>% filter(model == m) %>% drop_na()
    
    lmg_df = get_model_importance(regression_form, df=df, mixed_effects = F)
    colnames(lmg_df) = columnnames
    
    
    if (toggle_surp) {
      xlang_lmg_df = rbind(xlang_lmg_df, lmg_df %>% mutate(lang = l, model = m, orthogonalized = FALSE, surp = TRUE))
    } else{
      xlang_lmg_df = rbind(xlang_lmg_df, lmg_df %>% mutate(lang = l, model = m, orthogonalized = FALSE, surp = FALSE))
    }

  }
}

if (toggle_surp) {
  regression_form = "gaze_rt ~ orthogonal_surp + prev_orthogonal_surp + freq + prev_freq + orthogonal_len + prev_orthogonal_len"
} else {
  regression_form = "gaze_rt ~ orthogonal_pmi + prev_orthogonal_pmi + freq + prev_freq + orthogonal_len + prev_orthogonal_len"
}

for (l in langs) {
  
  print(paste0("Fitting orthogonalized model for ", l))
  
  df = read.csv(paste0("./merged_data/", l, ".csv")) %>%
    filter(freq > 0, prev_freq > 0, prev2_freq > 0) %>%
    mutate( freq = -log2(freq), prev_freq = -log2(prev_freq), prev2_freq = -log2(prev2_freq),
            pmi = freq - surp, 
            prev_pmi = prev_freq - prev_surp, 
            prev2_pmi = prev2_freq - prev2_surp)
  
  for (m in models) {
    
    df_eval = df %>% filter(model == m) %>% drop_na()
    
    lmg_df = get_model_importance(regression_form, df=df, mixed_effects = F)
    colnames(lmg_df) = columnnames
    
    if (toggle_surp) {
      xlang_lmg_df = rbind(xlang_lmg_df, lmg_df %>% mutate(lang = l, model = m, orthogonalized = TRUE, surp = TRUE))
    } else{
      xlang_lmg_df = rbind(xlang_lmg_df, lmg_df %>% mutate(lang = l, model = m, orthogonalized = TRUE, surp = FALSE))
    }

  }
}

if (toggle_surp) {
      xlang_lmg_df_surp = xlang_lmg_df
    } else {
      xlang_lmg_df_pmi = xlang_lmg_df
}

```

## Need to run the previous block for both values of toggle_surp before proceeding

```{r}

xlang_lmg_df = rbind(xlang_lmg_df_surp, xlang_lmg_df_pmi)

```


## Plot the LMG values

```{r}

coeff_plotting_df = xlang_lmg_df %>%
  gather(name, value, c(2:length(columnnames))) %>%
  group_by(lang, name, model, orthogonalized, surp) %>%
    summarise(m = mean(value),
              s = std.error(value),
              upper = m + s * 1.96,
              lower = m - s * 1.96) %>%
  ungroup() %>%
  filter((surp == TRUE & orthogonalized == TRUE) | (surp == TRUE & orthogonalized == FALSE) | (surp == FALSE & orthogonalized == FALSE) ) %>%
  mutate(position = case_when(
    #name %in% c("R2") ~ "0",
    name %in% c("frequency", "surp OR pmi", "length") ~ "0",
    name %in% c("previous frequency", "previous surp OR pmi", "previous length") ~ "1",
  )) %>%
  mutate(lmodel = case_when(
    orthogonalized == FALSE & surp == TRUE ~ "Surprisal",
    orthogonalized == FALSE & surp == FALSE ~ "PMI",
    orthogonalized == TRUE ~ "Orthogonalized",
  )) %>%
  mutate(predictor = case_when(
    name %in% c("frequency", "previous frequency") ~ "Frequency",
    name %in% c("surp OR pmi", "previous surp OR pmi") ~ "Context",
    name %in% c("length", "previous length") ~ "Length",
  )) %>%
  mutate(lmodel = factor(lmodel, levels = c("Surprisal", "PMI", "Orthogonalized"))) %>%
  mutate(predictor = factor(predictor, levels = c("Frequency", "Context", "Length"), labels = c("Frequency (Unigram Surprisal)", "Context (Contextual Surprisal / PMI / Orthogonalized Surprisal)", "Length / Orthogonalized Length"))) %>%
  mutate(position = factor(position, labels = c("w t", "w t-1"), levels = c("0", "1")))
```

```{r}
coeff_plotting_df %>%
    mutate(lang = factor(lang, levels = c("du", "en", "fi", "ge", "gr", "he", "it", "ko", "ru", "sp", "tr"),
       labels = c("Dutch", "English", "Finnish", "German", "Greek", "Hebrew", "Italian", "Korean", "Russian",
                  "Spanish", "Turkish"))) %>%
  ggplot(aes(x = position, y = m, fill = predictor)) +
    geom_bar(stat="identity", position = position_dodge(width = 0.9)) +
    geom_errorbar(aes(ymin=lower, ymax=upper), position = position_dodge(width = 0.9), width = 0.1) +
    geom_hline(yintercept = 0, color = "black") +
  ylab("Explained Variance (LMG)") +
  facet_grid(lmodel~lang, scales = "free_y") +
  scale_x_discrete(labels = c(bquote(w[t]), bquote(w[t-1]))) +
  #scale_color_manual(values = c("#225ea8", "#41b6c4", "#a1dab4")) +
  theme(
    text = element_text(family = "serif", size=11),
    axis.title.x = element_blank(),
    legend.position = "bottom"
  )

ggsave("./images/orthogonalized_comparison_lmg_length.pdf", width = 9, height = 5)

```

### Projecting Frequency onto the orthogonal complement of surprisal

```{r}

get_model_importance = function(form, df, mixed_effects, num_folds=10){
  
  folds <- cut(seq(1,nrow(df)),breaks=num_folds,labels=FALSE)
  
  metrics = data.frame()
  for(i in 1:num_folds){
    testIndexes = which(folds==i,arr.ind=TRUE)
    trainData = df[-testIndexes,]
    
    trainData = trainData %>% 
      mutate(freq = freq - mean(freq), prev_freq = prev_freq - mean(prev_freq), prev2_freq = prev2_freq - mean(prev2_freq),
           surp = surp - mean(surp), prev_surp = prev_surp - mean(prev_surp), prev2_surp = prev2_surp - mean(prev2_surp),
           pmi = pmi - mean(pmi), prev_pmi = prev_pmi - mean(prev_pmi), prev2_pmi = prev2_pmi - mean(prev2_pmi),
           len = len - mean(len), prev_len = prev_len - mean(prev_len), prev2_len = prev2_len - mean(prev2_len)) %>%
      mutate(orthogonal_freq = freq - (cov(trainData$surp, trainData$freq)/cov(trainData$surp, trainData$surp)) * surp,
              prev_orthogonal_freq = prev_freq - (cov(trainData$prev_surp, trainData$prev_freq)/cov(trainData$prev_surp, trainData$prev_surp)) * prev_surp,
              )

    if(mixed_effects){
      model = lmer(as.formula(form), data = trainData)
    } else {
      model = lm(as.formula(form), data = trainData)
    }
    
    res = calc.relimp(model, type = c("lmg"))
    metrics = rbind(metrics, c(R2=res$R2, res$lmg))
  }
  
  
  return(metrics)
}

```

```{r}



xlang_lmg_df = data.frame(matrix(ncol = 5, nrow = 0))
columnnames = c("intercept", "surp", "previous surp", "frequency", "previous frequency")
colnames(xlang_lmg_df) = columnnames

regression_form = "gaze_rt ~ surp + prev_surp + freq + prev_freq"

for (l in langs) {
  
  print(paste0("Fitting standard model for ", l))
  
  df = read.csv(paste0("./merged_data/", l, ".csv")) %>%
    filter(freq > 0, prev_freq > 0, prev2_freq > 0) %>%
    mutate( freq = -log2(freq), prev_freq = -log2(prev_freq), prev2_freq = -log2(prev2_freq),
            pmi = freq - surp, 
            prev_pmi = prev_freq - prev_surp, 
            prev2_pmi = prev2_freq - prev2_surp)
  
  for (m in models) {
    
    df_eval = df %>% filter(model == m) %>% drop_na()
    
    lmg_df = get_model_importance(regression_form, df=df, mixed_effects = F)
    colnames(lmg_df) = columnnames
    
    xlang_lmg_df = rbind(xlang_lmg_df, lmg_df %>% mutate(lang = l, model = m, orthogonalized = FALSE))

  }
}

regression_form = "gaze_rt ~ surp + prev_surp + orthogonal_freq + prev_orthogonal_freq"



for (l in langs) {
  
  print(paste0("Fitting orthogonalized model for ", l))
  
  df = read.csv(paste0("./merged_data/", l, ".csv")) %>%
    filter(freq > 0, prev_freq > 0, prev2_freq > 0) %>%
    mutate( freq = -log2(freq), prev_freq = -log2(prev_freq), prev2_freq = -log2(prev2_freq),
            pmi = freq - surp, 
            prev_pmi = prev_freq - prev_surp, 
            prev2_pmi = prev2_freq - prev2_surp)
  
  for (m in models) {
    
    df_eval = df %>% filter(model == m) %>% drop_na()

    
    lmg_df = get_model_importance(regression_form, df=df, mixed_effects = F)
    colnames(lmg_df) = columnnames
    
    xlang_lmg_df = rbind(xlang_lmg_df, lmg_df %>% mutate(lang = l, model = m, orthogonalized = TRUE))
      

  }
}

```

## Plot the LMG values

```{r}

coeff_plotting_df = xlang_lmg_df %>%
  gather(name, value, c(2:length(columnnames))) %>%
  group_by(lang, name, model, orthogonalized) %>%
    summarise(m = mean(value),
              s = std.error(value),
              upper = m + s * 1.96,
              lower = m - s * 1.96) %>%
  ungroup() %>%
  mutate(position = case_when(
    #name %in% c("R2") ~ "0",
    name %in% c("frequency", "surp") ~ "0",
    name %in% c("previous frequency", "previous surp") ~ "1",
  )) %>%
  mutate(lmodel = case_when(
    orthogonalized == FALSE ~ "Frequency",
    orthogonalized == TRUE ~ "Orth. Frequency",
  )) %>%
  mutate(predictor = case_when(
    name %in% c("frequency", "previous frequency") ~ "Frequency",
    name %in% c("surp", "previous surp") ~ "Surprisal",
    #name %in% c("R2") ~ "R2",
  )) %>%
  mutate(lmodel = factor(lmodel, levels = c("Frequency", "Orth. Frequency"))) %>%
  mutate(predictor = factor(predictor, levels = c("Frequency", "Surprisal"), labels = c("Frequency (Frequency / Orthogonalized Frequency)", "Context (Surprisal)"))) %>%
  mutate(position = factor(position, labels = c("w t", "w t-1"), levels = c("0", "1")))
```

```{r}
coeff_plotting_df %>%
    mutate(lang = factor(lang, levels = c("du", "en", "fi", "ge", "gr", "he", "it", "ko", "ru", "sp", "tr"),
       labels = c("Dutch", "English", "Finnish", "German", "Greek", "Hebrew", "Italian", "Korean", "Russian",
                  "Spanish", "Turkish"))) %>%
  ggplot(aes(x = position, y = m, fill = predictor)) +
    geom_bar(stat="identity", position = position_dodge(width = 0.9)) +
    geom_errorbar(aes(ymin=lower, ymax=upper), position = position_dodge(width = 0.9), width = 0.1) +
    geom_hline(yintercept = 0, color = "black") +
  ylab("Explained Variance (LMG)") +
  facet_grid(lmodel~lang, scales = "free_y") +
  scale_x_discrete(labels = c(bquote(w[t]), bquote(w[t-1]))) +
  #scale_color_manual(values = c("#225ea8", "#41b6c4", "#a1dab4")) +
  theme(
    text = element_text(family = "serif", size=14),
    axis.title.x = element_blank(),
    legend.position = "bottom"
  )

ggsave("./images/orthogonalized_frequency_comparison_lmg.pdf", width = 9, height = 5)

```



## EXPERIMENT 2: GAMs to study effect of each predictor
## Compute DLL for Each Language

```{r}

model_cross_val = function(form, id, df, d_var, mixed_effects, psychometric, lang, num_folds=10, linear=FALSE){
  
  folds <- cut(seq(1,nrow(df)),breaks=num_folds,labels=FALSE)

  estimates <- c()
  models <- c()
  for(i in 1:num_folds){
    testIndexes = which(folds==i,arr.ind=TRUE)
    testData = df[testIndexes,]
    trainData = df[-testIndexes,]
    
    trainData = trainData %>% mutate(orthogonal_surp = surp - (cov(trainData$freq, trainData$surp)/cov(trainData$freq, trainData$freq)) * freq,
                                     prev_orthogonal_surp = prev_surp - (cov(trainData$prev_freq, trainData$prev_surp)/cov(trainData$prev_freq, trainData$prev_freq)) * prev_freq,
                                     orthogonal_pmi = pmi - (cov(trainData$freq, trainData$pmi)/cov(trainData$freq, trainData$freq)) * freq,
                                     prev_orthogonal_pmi = prev_pmi - (cov(trainData$prev_freq, trainData$prev_pmi)/cov(trainData$prev_freq, trainData$prev_freq)) * prev_freq)
    
    testData = testData %>% mutate(orthogonal_surp = surp - (cov(trainData$freq, trainData$surp)/cov(trainData$freq, trainData$freq)) * freq,
                                     prev_orthogonal_surp = prev_surp - (cov(trainData$prev_freq, trainData$prev_surp)/cov(trainData$prev_freq, trainData$prev_freq)) * prev_freq,
                                     orthogonal_pmi = pmi - (cov(trainData$freq, trainData$pmi)/cov(trainData$freq, trainData$freq)) * freq,
                                     prev_orthogonal_pmi = prev_pmi - (cov(trainData$prev_freq, trainData$prev_pmi)/cov(trainData$prev_freq, trainData$prev_freq)) * prev_freq)
    
    if (linear) {
      model = lm(as.formula(form), data = trainData)
    } else {
      if(id == "baseline") {
        model = lm(as.formula(form), data = trainData)
      } 
      else{
        model = gam(as.formula(form), data = trainData)
      }
    }
    
    # Save models to use later for looking at the effect
    # Don't save the ensemble models
    if(id %in% c("freq", "pmi", "surp", "pmi+freq", "surp+freq", "all", "baseline", "opmi", "osurp", "opmi+freq", "osurp+freq")) {
      model_id = paste(lang, psychometric, id, i, sep="-")
      saveRDS(model, file=paste("./models/", model_id,".rds", sep=""))
    }
    
    stdev = sigma(model)
    densities <- log(dnorm(testData[[d_var]],mean=predict(model, newdata=testData),sd=stdev))
    estimates <- c(estimates, densities)
  }

  return(estimates)
}

```


```{r}

regression_names = c("surp", "freq", "pmi", "pmi+freq", "surp+freq", "baseline", "opmi", "osurp", "opmi+freq", "osurp+freq")

dll_raw_df = data.frame()

for (lang in langs) {

  print(paste0("Fitting model for ", lang))
  
  df = read.csv(paste0("./merged_data/", lang, ".csv")) %>%
    filter(freq > 0, prev_freq > 0, prev2_freq > 0) %>%
    mutate( freq = -log2(freq), prev_freq = -log2(prev_freq), prev2_freq = -log2(prev2_freq),
            pmi = freq - surp, 
            prev_pmi = prev_freq - prev_surp, 
            prev2_pmi = prev2_freq - prev2_surp)
  
  for (m in models) {
    
    df_eval = df %>% filter(model == m) %>% drop_na()
  
    for (psychometric in psychometrics) {
      
      regression_forms = c(
        #paste0(psychometric, " ~ surp + prev_surp + prev2_surp + freq + prev_freq + prev2_freq + len + prev_len + prev2_len"), # SURPRISAL
        #paste(psychometric, "~ s(surp, bs = 'cr', k = 6) + s(prev_surp, bs = 'cr', k = 6) + s(len, bs = 'cr', k = 6) + s(prev_len, bs = 'cr', k = 6)"), #SURPRISAL
        paste(psychometric, "~ s(surp, bs = 'cr', k = 6) + s(prev_surp, bs = 'cr', k = 6)"), #SURPRISAL
        
        #paste0(psychometric, " ~ freq + prev_freq + prev2_freq + len + prev_len + prev2_len"), # FREQUENCY
        paste(psychometric, "~ s(freq, bs = 'cr', k = 6) + s(prev_freq, bs = 'cr', k = 6)"), #FREQUENCY
        
        #paste0(psychometric, " ~ pmi + prev_pmi + prev2_pmi + freq + prev_freq + prev2_freq + len + prev_len + prev2_len"), # PMI
        paste(psychometric, "~ s(pmi, bs = 'cr', k = 6) + s(prev_pmi, bs = 'cr', k = 6)"), #PMI
        
        paste(psychometric, "~ s(pmi, bs = 'cr', k = 6) + s(prev_pmi, bs = 'cr', k = 6) + s(freq, bs = 'cr', k = 6) + s(prev_freq, bs = 'cr', k = 6)"), #PMI+FREQ
        
        paste(psychometric, "~ s(surp, bs = 'cr', k = 6) + s(prev_surp, bs = 'cr', k = 6) + s(freq, bs = 'cr', k = 6) + s(prev_freq, bs = 'cr', k = 6)"), #SURP+FREQ


#        paste(psychometric, "~ s(pmi, bs = 'cr', k = 6) + s(prev_pmi, bs = 'cr', k = 6) + s(freq, bs = 'cr', k = 6) + s(prev_freq, bs = 'cr', k = 6) + s(surp, bs = 'cr', k = 6) + s(prev_surp, bs = 'cr', k = 6)+  te(len, bs = 'cr') + te(prev_len, bs = 'cr')"), #ALL

        #paste0(psychometric, " ~ freq + prev_freq + prev2_freq + len + prev_len + prev2_len") # BASELINE
        #paste(psychometric, "~ s(len, bs = 'cr', k = 6) + s(prev_len, bs = 'cr', k = 6)"), #BASELINE
        paste(psychometric, "~ 1"), #BASELINE

        paste(psychometric, "~ s(orthogonal_pmi, bs = 'cr', k = 6) + s(prev_orthogonal_pmi, bs = 'cr', k = 6)"), #OPMI

        paste(psychometric, "~ s(orthogonal_surp, bs = 'cr', k = 6) + s(prev_orthogonal_surp, bs = 'cr', k = 6)"), #OSURP

        paste(psychometric, "~ s(orthogonal_pmi, bs = 'cr', k = 6) + s(prev_orthogonal_pmi, bs = 'cr', k = 6) + s(freq, bs = 'cr', k = 6) + s(prev_freq, bs = 'cr', k = 6)"), #OPMI+FREQ

        paste(psychometric, "~ s(orthogonal_surp, bs = 'cr', k = 6) + s(prev_orthogonal_surp, bs = 'cr', k = 6) + s(freq, bs = 'cr', k = 6) + s(prev_freq, bs = 'cr', k = 6)") #OSURP+FREQ
      )
      
      loglik_df = data.frame(names=regression_names, forms=regression_forms) %>%
        mutate(logliks = map2(regression_forms, regression_names, model_cross_val, df=df_eval, d_var=psychometric, mixed_effects=F, psychometric=psychometric, lang=lang )) %>%
        dplyr::select(-forms)
      
      loglik_df = loglik_df %>% unnest(cols = c(logliks)) %>% mutate(lang = lang, psychometric = psychometric, model = m)
      dll_raw_df = rbind(dll_raw_df, loglik_df)
      
    }
  }
}


```

## Data for each language individually

```{r}
#comps = c("surp", "freq", "pmi")
comps = c("surp", "freq", "pmi", "pmi+freq", "surp+freq", "osurp", "osurp+freq")


dll_xlang_surp_df = data.frame()
for(l in langs){
  
  print(paste0("Tests for ", l)) 
  
  for (ps in psychometrics){
    for(c in comps){
      for(m in models) {

        if(c != "baseline") {
          target_df = dll_raw_df %>% filter(psychometric == ps, names == c, lang == l, model == m)
          baseline_df = dll_raw_df %>% filter(psychometric == ps, names == "baseline", lang == l, model == m)
          dll = target_df$logliks - baseline_df$logliks
          dll = dll[!is.na(dll)]
          ptest = perm.test(dll, num.sim = 500)
          dll_df = data.frame(comp = c, mean = mean(dll), upper = mean(dll) + (1.96 * std.error(dll)),
                              lower = mean(dll) - (1.96 * std.error(dll)), ptest_pval = ptest$p.value,
                              lang = l, psychometric = ps, model = m)
          dll_xlang_surp_df = rbind(dll_xlang_surp_df, dll_df)
        }
      }
    }
  }
}

```

## Data for languages as a whole

```{r}

dll_agg_surp_df = data.frame()
for (ps in psychometrics){
  for(c in comps){
    for(m in models) {

        if(c != "baseline") {
          target_df = dll_raw_df %>% filter(psychometric == ps, names == c, model == m)
          baseline_df = dll_raw_df %>% filter(psychometric == ps, names == "baseline", model == m)
          dll = target_df$logliks - baseline_df$logliks
          dll = dll[!is.na(dll)]
          ptest = perm.test(dll, num.sim = 500)
          dll_df = data.frame(comp = c, mean = mean(dll), upper = mean(dll) + (1.96 * std.error(dll)),
                              lower = mean(dll) - (1.96 * std.error(dll)), ptest_pval = ptest$p.value,
                              lang = "All", psychometric = ps, model = m)
          dll_agg_surp_df = rbind(dll_agg_surp_df, dll_df)
        }
    }
  }
}


```


### Plot all the languages

```{r}
# Merge the dataframes
dll_xlang_surp_df %>%
  rename(target = comp) %>%
  mutate(sig = case_when( ptest_pval >= 0.05 ~ " ",
                          ptest_pval < 0.05 & ptest_pval >= 0.01  ~ "*",
                          ptest_pval < 0.01 & ptest_pval >= 0.001  ~ "**",
                          ptest_pval < 0.001  ~ "***")) %>%
  mutate(psychometric = case_when(psychometric == "firstfix_rt" ~ "First Fixation", 
                                  psychometric == "gaze_rt" ~ "Gaze Duration", 
                                  psychometric == "total_rt" ~ "Total Fixation")) %>%
  mutate(model = case_when(model == "mgpt_lc" ~ "mGPT",
                           model == "monot_all" ~ "monoT")) %>%
  mutate(model = factor(model, levels = c("mGPT", "monoT"))) %>%
  #mutate(target = factor(target, levels = c("freq", "surp", "pmi", "osurp", "surp+freq", "pmi+freq", "osurp+freq"), labels = c("Frequency", "Surprisal", "PMI", "Orthogonalized\nSurprisal/PMI", "Surprisal & Frequency", "PMI & Frequency", "Orthogonalized\nSurp./PMI & Freq."))) %>%
  mutate(target = factor(target, levels = c("freq", "surp", "pmi", "osurp", "surp+freq", "pmi+freq", "osurp+freq"), labels = c("Freq.", "Surp.", "PMI", "Orthog.\nSurp.", "Surp.\n& Freq.", "PMI\n& Freq.", "Orthog.\nSurp.\n& Freq."))) %>%
  
   
  ggplot(aes(x = target, y = mean, color = lang)) +
    #scale_x_discrete(labels =c("freq", "surp", "pmi", "orthogonalized\nsurprisal", "surp+freq", "pmi+freq", "orthogonalized\nsurprisal+freq")) + 
    geom_hline(yintercept=0, color="black", linetype="dashed", alpha =0.5) +
  
    #geom_point(position = position_dodge(width = 0.1), size = 2) +
    geom_errorbar(aes(ymin=lower, ymax=upper), width = 0.1, position = position_dodge(width = 0.8 ), alpha=0.3) +
    
    geom_label(aes(label=lang, group=lang), size=2.5, position = position_dodge(width = 0.8), label.padding=unit(0.5, "mm")) +
    #geom_text_repel(aes(label=lang, group=target), size=3, label.padding=unit(0.7, "mm"), min.segment.length = unit(0, 'lines')) +
  
    #geom_text(aes(y = 0.08, label = sig), size = 3, show.legend = FALSE) +
    ylab("Delta Log Likelihood \n (average  per word)") + 
    xlab("") +
    facet_grid(.~target, scales="free", space = "free") +
    labs(color = "") +
    guides(color=guide_legend(ncol=2)) +
    #scale_x_discrete(labels = c(bquote(w[t-2]), bquote(w[t-1]), bquote(w[t]))) +
    #scale_color_manual(values = c("#a1dab4", "#41b6c4", "#225ea8")) +
    #ggtitle("Contribution of Surprisal to ΔLL") +
  theme(
    strip.background = element_blank(),
    strip.text.x = element_blank(),
    text = element_text(family="serif"),
    legend.position = "none",
    axis.title.x = element_blank(),
    panel.spacing = unit(0.1, "lines"),
    axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5)#,
    #panel.border = element_rect(color = "grey", fill =  NA, size = 0.5)
  )
ggsave("./images/results_agg_all_langs.pdf", device = "pdf", width = 6, height = 3)


```


### Plot the cross language averages

```{r}

# Merge the dataframes
dll_agg_surp_df %>%
  rename(target = comp) %>%
  mutate(sig = case_when( ptest_pval >= 0.05 ~ " ",
                          ptest_pval < 0.05 & ptest_pval >= 0.01  ~ "*",
                          ptest_pval < 0.01 & ptest_pval >= 0.001  ~ "**",
                          ptest_pval < 0.001  ~ "***")) %>%
  mutate(psychometric = case_when(psychometric == "firstfix_rt" ~ "First Fixation", 
                                  psychometric == "gaze_rt" ~ "Gaze Duration", 
                                  psychometric == "total_rt" ~ "Total Fixation")) %>%
  mutate(model = case_when(model == "mgpt_lc" ~ "mGPT",
                           model == "monot_all" ~ "monoT")) %>%
  mutate(model = factor(model, levels = c("mGPT", "monoT"))) %>%
  mutate(target = factor(target, levels = c("freq", "surp", "pmi", "osurp", "surp+freq", "pmi+freq", "osurp+freq"), labels = c("frequency", "surprisal", "pmi", "orthogonalized surprisal/pmi", "surprisal and frequency", "pmi and frequency", "orthogonalized surprisal/pmi and frequency"))) %>%
  ggplot(aes(x = target, y = mean, color = target)) + scale_x_discrete(labels =c("freq", "surp", "pmi", "osurp", "surp+freq", "pmi+freq", "osurp+freq")) + 
    geom_hline(yintercept=0, color="black", linetype="dashed", alpha =0.5) +
    geom_point(position = position_dodge(width = 0.6), size = 3) +
    #geom_text(aes(y = 0.08, label = sig), size = 3, show.legend = FALSE) +
    geom_errorbar(aes(ymin=lower, ymax=upper), width = 0.1, position = position_dodge(width = 0.6)) +
    ylab("Delta Log Liklihood \n (average  per word)") + 
    xlab("") +
    facet_grid(.~psychometric) +
    labs(color = "RT Predictor") +
    #scale_x_discrete(labels = c(bquote(w[t-2]), bquote(w[t-1]), bquote(w[t]))) +
    #scale_color_manual(values = c("#a1dab4", "#41b6c4", "#225ea8")) +
    #ggtitle("Contribution of Surprisal to ΔLL") +
  theme(
    text = element_text(family="serif"),
    legend.position = "right",
    axis.title.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)#,
    #panel.border = element_rect(color = "grey", fill =  NA, size = 0.5)
  )

ggsave("./images/results_agg.pdf", device = "pdf", width = 8.5, height = 4)


```


### Plot everything in a combined figure


```{r}

# Merge the dataframes
dll_surp_plotting_df = rbind(dll_xlang_surp_df, dll_agg_surp_df)
options(scipen=999)

dll_surp_plotting_df %>%
  mutate(lang = factor(lang, levels = c("All", "du", "en", "fi", "ge", "gr", "he", "it", "ko", "ru", "sp", "tr"),
       labels = c("All", "Dutch", "English", "Finnish", "German", "Greek", "Hebrew", "Italian", "Korean", "Russian", "Spanish", "Turkish"))) %>%
  filter(model %in% c("mgpt_lc", "monot_all")) %>%
  rename(target = comp) %>%
  mutate(sig = case_when( ptest_pval >= 0.05 ~ " ",
                          ptest_pval < 0.05 & ptest_pval >= 0.01  ~ "*",
                          ptest_pval < 0.01 & ptest_pval >= 0.001  ~ "**",
                          ptest_pval < 0.001  ~ "***")) %>%
  mutate(psychometric = case_when(psychometric == "firstfix_rt" ~ "First Fixation", 
                                  psychometric == "gaze_rt" ~ "Gaze Duration", 
                                  psychometric == "total_rt" ~ "Total Fixation")) %>%
  mutate(model = case_when(model == "mgpt_lc" ~ "mGPT",
                           model == "monot_all" ~ "monoT")) %>%
  mutate(model = factor(model, levels = c("mGPT", "monoT"))) %>%
  mutate(target = factor(target,levels = c("freq", "surp", "pmi", "osurp", "surp+freq", "pmi+freq", "osurp+freq"))) %>%
  ggplot(aes(x = target, y = mean, color = target)) +
    geom_hline(yintercept=0, color="black", linetype="dashed", alpha =0.5) +
    geom_point(position = position_dodge(width = 0.6), size = 2) +
    #geom_text(aes(y = 0.08, label = sig), size = 3, show.legend = FALSE) +
    geom_errorbar(aes(ymin=lower, ymax=upper), width = 0.1, position = position_dodge(width = 0.6)) +
    ylab("Delta Log Liklihood (average  per word)") + 
    xlab("") +
    facet_grid(psychometric~lang) +
    labs(color = "Eye Movement Measure") +
    #scale_x_discrete(labels = c(bquote(w[t-2]), bquote(w[t-1]), bquote(w[t]))) +
    #scale_color_manual(values = c("#a1dab4", "#41b6c4", "#225ea8")) +
    #ggtitle("Contribution of Surprisal to ΔLL") +
  theme(
    text = element_text(family="serif"),
    legend.position = "bottom",
    axis.title.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)#,
    #panel.border = element_rect(color = "grey", fill =  NA, size = 0.5)
  )

ggsave("./images/results_gd.pdf", device = "pdf", width = 11.5, height = 6)


```


